Windowed residual-minimization model reduction for dynamical systems

### 2019-10-03 main.tex
working on intro

The LSPG approach sequentially performs residual minimization over a single
time-step, and as such is not associated with residual minimization over the
entire time interval. \KTC{This is what ST-LSPG does. Why do we ignore it? the
full dynamical system} Indeed, LSPG is not equipped with \textit{a priori}
guarantees of stability or convergence, even for linear time-invariant
systems~\cite{bui_thesis}. Further, the performance of LSPG depends on the
time step in two distinct ways: changing the time-step size modifies both the
time window over which LSPG minimizes the residual \textit{as well as} the
time-discretization error of the full-order model on which LSPG is based. As a
result, LSPG often yields the smallest error for an intermediate value of the
time step (see, e.g., Ref.~\cite[Figure 9]{carlberg_lspg_v_galerkin}). In
particular, as the time step decreases, the accuracy of LSPG approaches that
of Galerkin (which is often poor)\footnote{This occurs because LSPG projection
is equivalent to Galerkin projection in the limit $\Delta t \rightarrow 0$.};
as the time step increases, the time discretization error of the full-order
model (on which the LSPG ROM is based) increases.  LSPG projection additionally is
additionally sensitive to the time discretization scheme. When used with
explicit temporal discretization, for instance, LSPG again reverts to
Galerkin. To gain the enhanced stability associated with LSPG, the underlying
time-discretization scheme must be implicit with an appropriate time-step.
This can increase the cost of the reduced-order model, especially in the case
where the optimal time-step required by LSPG is small.

@@

The paper proceeds as follows: Section~\ref{sec:math}
outlines the mathematical setting for the full-order model, along with the Galerkin, LSPG, and
space--time LSPG ROMs; Section~\ref{sec:tclspg} outlines the proposed \methodAcronym\
formulation; Section~\ref{sec:numerical_techniques}
outlines numerical techniques for solving \methodAcronymROMs, including both direct and
indirect methods; Section~\ref{sec:analysis} provides
equivalence conditions and error analysis for \methodAcronymROMs;
Section~\ref{sec:numerical_experiments} presents numerical experiments;
Section~\ref{sec:conclude} provides conclusions and perspectives.
We denote vector-valued functions with stylized bold text (i.e., $\boldsymbol
x$), vectors with standard bold text (i.e., $\mathbf{x}$), 
matrices with capital bold letters (i.e., $\mathbf{X} \equiv \begin{bmatrix}
\mathbf{x}_1 & \cdots & \mathbf{x}_r\end{bmatrix}$), and spaces with
calligraphic fonts (i.e., $\mathcal{X}$).
%Model reduction of nonlinear dynamical systems is a growing research area.
%The least-squares Petrov-Galerkin (LSPG) approach has gained popularity as a model reduction technique for nonlinear dynamical systems. Derived at the fully discrete level, the LSPG approach solves a least-squares problem that minimizes the fully discrete residual of the reduced-order model at each time-step. LSPG has been shown to be an effective formulation for nonlinear model reduction on complex problems of interest.
%\begin{comment} 
\section{Mathematical setting}\label{sec:math}
We consider the system of nonlinear ordinary differential equations (ODEs),
\begin{equation}\label{eq:FOM}
 \frac{d\stateFOM}{dt} = \velocity(\stateFOM,t), \qquad \stateFOM(0) =
	\stateFOMIC,\qquad t \in [0,T]
\end{equation}
where 
$T \in \mathbb{R}^+$ denotes the final time, $\stateFOM \in \RR{\fomdim}
	\otimes \timeSpace $ with $\stateFOM: [0,T] \rightarrow  \RR{\fomdim}$
	denotes the state implicitly defined as the solution to initial value
	problem~\eqref{eq:FOM}, $\stateFOMIC \in \mathbb{R}^{\fomdim}$ denotes the initial condition,
	and $\velocity: \mathbb{R}^{\fomdim} \times [0,T] \rightarrow
	\mathbb{R}^{\fomdim}$ with $(\stateyDiscrete,t) \mapsto
	\velocity(\stateyDiscrete,t)$ denotes the velocity, which is possibly
	nonlinear in its first argument. We refer to Eq.~\eqref{eq:FOM} 
	as the ``full-order model" (FOM) ODE.

@@

\subsection{Projection-based reduced-order models}
Projection-based ROMs generate approximate solutions to the FOM
	ODE~\eqref{eq:FOM} by approximating the state in a low-dimensional trial
	subspace. For dynamical 
	systems, two distinct types of space--time trial subspaces are commonly used: 
\begin{enumerate} 
	\item \textit{\spatialAcronym\ trial subspaces}. These 
	trial subspaces reduce only the spatial dimension of the FOM, and are commonly used
		in model-reduction methods based on spatial projection followed by time
		integration (e.g., Galerkin projection, LSPG projection)
	\item \textit{\spaceTimeAcronym\ trial subspaces}. These 
	trial subspaces reduce both the spatial and temporal dimensions of the FOM,
		and are commonly employed 
		employed in model-reduction methods based on space--time projection (e.g.,
		space--time LSPG projection).
\end{enumerate}
 We now describe these two types of trial subspaces and their application to 
	Galerkin, LSPG, and space--time LSPG projection. 
	\KTC{HERE}

\subsection{\spatialAcronym\ trial subspaces}
\spatialAcronym\ trial subspaces approximate the FOM ODE solution
	$\stateFOM$ with an approximation $\approxstate$ that resides in an
	affine space--time trial subspace of dimension $K\ll\fomdim$, i.e., 
\begin{align}\label{eq:spatial_subspace}
& \stspaceS \defeq \trialspace \otimes \timeSpace +
	\stateIntercept\otimes\onesFunction,\\ 
& \trialspace \defeq \text{Range}(\basismat),  \label{eq:spatial_subspace2}  \\
&\basismat \defeq \begin{bmatrix}  \basisvec_1  & \cdots &  \basisvec_K \end{bmatrix}, \label{eq:spatial_subspace3} 
\end{align}
where 
	$\timeSpace$ denotes the set of real-valued functions acting on the time
	domain
	$[0,T]$, $\onesFunction\in\timeSpace$ is defined as $\onesFunction:[0,T]\rightarrow\{1\}$,
	$\basismat \in \RRStar{\romdim}{\fomdim}$ denotes the spatial trial subspace with
	$\RRStar{\romdim}{\fomdim}$ denoting the compact Stiefel manifold (i.e.,  $
	\RRStar{\romdim}{\fomdim}\defeq
	\{ \mathbf{X} \in \RR{\fomdim
	\times \romdim}\, \big|\, \mathbf{X}^T \mathbf{X} = \mathbf{I} \}$),
	$\trialspace\subseteq\RR{\fomdim}$ denotes the spatial trial subspace, and
	$\stateIntercept \in \mathbb{R}^{\fomdim}$ denotes the reference state that
	is often taken to be the initial condition (i.e., $\stateIntercept = \stateFOMIC$). 
%where $\basisvec_i \in \mathbb{R}^N$ are orthonormal basis vectors and $\stateIntercept \in \mathbb{R}^{\fomdim}$ the ``intercept''.
%We refer to $\basismat \in \RRStar{\fomdim \times \romdim}$, with $\RRStar{p \times q}$ denoting the set of orthogonal $q$ frames in $p$, as the spatial trial basis, and $\trialspace$ as the spatial trial space. 
The basis vectors $\basisvec_i$, $i=1,\ldots,K$ are typically constructed
using state snapshots, e.g., via
proper orthogonal decomposition (POD)~\cite{berkooz_turbulence_pod} or the reduced-basis method~\cite{rb_1,rb_2,rb_3,NgocCuong2005,Rozza2008}. 
At any time instance $t\in[0,T]$, ROMs that employ the  \spatialAcronym\ trial subspace approximate the state as
\begin{equation}\label{eq:affine_trialspace}
\stateFOMArg{}{t} \approx \approxstate(t) = \basisspace \genstateArg{}{t} + \stateIntercept,
\end{equation}
where $\genstate \in \RR{\romdim} \otimes \timeSpace$ denotes the generalized
coordinates. Substituting the approximation~\eqref{eq:affine_trialspace} in
Eq.~\eqref{eq:FOM} and performing orthogonal
$\elltwo$-projection of the initial condition onto the trial subspace yields
the overdetermined system of ODEs
\begin{equation}\label{eq:g_truncation}
\basisspace \frac{d\genstate }{dt} = \velocity(\basisspace
\genstate + \stateIntercept,t ), \qquad \genstate(0) = \genstateIC, \qquad t \in [0,T].
\end{equation}
Because Eq.~\eqref{eq:g_truncation} is overdetermined, a solution may not
exist. Typically, either \textit{Galerkin projection} or \textit{least-squares
Petrov--Galerkin projection} is employed to reduce the number of equations
such that a unique solution exists. We now describe these two methods.

@@

\frac{d\genstateGalerkin}{dt} = \massArgnt{n}^{-1} \basisspace^T \stweightingMatArg{n} \velocity(\basisspace
\genstateGalerkin + \stateIntercept,t), \qquad \genstateGalerkin(0) = \genstateIC, \qquad t \in [0,T],

@@

\frac{d\genstateGalerkin}{dt} = \underset{ \genstateyDiscrete \in \RR{\romdim}

@@

	\velocity(\basisspace \genstateGalerkin + \stateIntercept,t)

>>>>

The LSPG approach has several shortcomings. First, the performance of LSPG depends on the
time discretization. Changing the time-step size modifies both the
time window over which LSPG minimizes the residual \textit{as well as} the
time-discretization error of the full-order model on which LSPG is based. As a
result, LSPG often yields the smallest error for an intermediate value of the
time step (see, e.g., Ref.~\cite[Figure 9]{carlberg_lspg_v_galerkin}). In
particular, as the time step decreases, the accuracy of LSPG approaches that
of Galerkin (which is often poor)\footnote{This occurs because LSPG projection
is equivalent to Galerkin projection in the limit $\Delta t \rightarrow 0$.}. 
%The accuracy of LSPG additionally degrades if the time-step size becomes too large. It is unclear if this is due to the discretization error associated with enlarging the time-step size, or rather if it is due to the size of the window the residual is being minimized over. LSPG projection additionally is
%additionally sensitive to the time discretization scheme. When used with
%explicit temporal discretization, for instance, LSPG again reverts to
%Galerkin. To gain the enhanced stability associated with LSPG, the underlying
%time-discretization scheme must be implicit with an appropriate time-step.
%%This can increase the cost of the reduced-order model, especially in the case
%where the optimal time-step required by LSPG is small. 
Second, as the LSPG approach sequentially performs residual minimization over
time-steps, is not equipped with any notion of optimality over the entire 
time domain of interest. As a result, LSPG is not equipped with \textit{a priori}
guarantees of stability or convergence, even for linear time-invariant
systems~\cite{bui_thesis}.

Researchers have pursued the development of space--time residual minimization ROMs~\cite{choi_stlspg,constantine_strom,URBAN2012203,Yano2014ASC,bui_thesis} to address this last issue. In existing space--time methods, the spatial \textit{and} temporal dependence of the state is represented in a low-dimensional space--time trial space. Space--time residual minimization approaches then comprise computing a space--time solution within this low-dimensional space--time trial space that minimizes the space--time norm of the full-order model. Space--time methods are thus equipped with a notion of optimality over the entire time domain of interest and, as a result, contain more favorable error bounds than the Galerkin and LSPG approaches.  Several drawbacks of existing space--time ROMs are that they (1) require minimizing the entire space--time residual (which can be a computationally intensive process) and (2) require the specification of a finite-dimensional temporal basis (computed, e.g., via POD applied to space--time snapshots). This latter issue is extremely relevant to the present work.  
%These space--time In these space--time ROMs, the temporal dependence of the state 
%is represented via a finite dimensional basis representation. 
%Space--time residual minimization methods can be prohibitively expensive, however, as they involve minimizing the entire space--time residual. Further, existing space--time approaches require a finite-dimensional space--time basis.

%This work seeks to overcome the practical challenges associated with existing
%residual minimization approaches for reduced-order models. To this end, we
%propose the \methodName\ method for model reduction, which minimizes the FOM ODE residual
%over a sequence of time windows. The key novelties of the work are two fold.
%First, the residual minimization problem is formulated to minimize the
%\textit{time-continuous} residual. By working from a time-continuous viewpoint, the
%present formulation decouples the underlying temporal discretization scheme
%from the residual minimization problem; thus addressing a key deficiency of
%LSPG \KTC{This is also true for ST-LSPG and maybe other discretization-based
%methods? Have to check}. The time-continuous viewpoint additionally exposes
%two different methods that
%can be used to solve the residual-minimization problems, namely
%a \textit{discretize-then-optimize} (i.e., direct) method and
%an \textit{optimize-then-discretize} (i.e., indirect) method. The second novelty
%is that the residual is minimized over \textit{time windows} rather than over
%a single time step or the entire time domain. This equips the method with
%additionaly flexibility that enables it
%to explore more fine-grained tradeoffs in computational cost and error. We
%demonstrate that this method can be viewed as a generalization of existing
%methods, as particular instances of this formulation recover Galerkin, LSPG,
%and space--time LSPG projection. Figure~\ref{fig:flowchart} depicts how the
%proposed method is related to existing approaches.

% \KTC{This is what ST-LSPG does. Why do we ignore it? the
%full dynamical system} Indeed, LSPG is not equipped with \textit{a priori}
%guarantees of stability or convergence, even for linear time-invariant
%systems~\cite{bui_thesis}. Further, the performance of LSPG depends on the
%time step in two distinct ways: changing the time-step size modifies both the
%time window over which LSPG minimizes the residual \textit{as well as} the
%time-discretization error of the full-order model on which LSPG is based. As a
%result, LSPG often yields the smallest error for an intermediate value of the
%time step (see, e.g., Ref.~\cite[Figure 9]{carlberg_lspg_v_galerkin}). In
%particular, as the time step decreases, the accuracy of LSPG approaches that
%of Galerkin (which is often poor)\footnote{This occurs because LSPG projection
%is equivalent to Galerkin projection in the limit $\Delta t \rightarrow 0$.};
%as the time step increases, the time discretization error of the full-order
%model (on which the LSPG ROM is based) increases.  LSPG projection additionally is
%additionally sensitive to the time discretization scheme. When used with
%explicit temporal discretization, for instance, LSPG again reverts to
%Galerkin. To gain the enhanced stability associated with LSPG, the underlying
%time-discretization scheme must be implicit with an appropriate time-step.
%This can increase the cost of the reduced-order model, especially in the case
%where the optimal time-step required by LSPG is small.

@@

The paper proceeds as follows: Section~\ref{sec:math}
outlines the mathematical setting for the full-order model, along with the Galerkin, LSPG, and
space--time LSPG ROMs; Section~\ref{sec:tclspg} outlines the proposed \methodAcronym\
formulation; Section~\ref{sec:numerical_techniques}
outlines numerical techniques for solving \methodAcronymROMs, including both direct and
indirect methods; Section~\ref{sec:analysis} provides
equivalence conditions and error analysis for \methodAcronymROMs;
Section~\ref{sec:numerical_experiments} presents numerical experiments;
Section~\ref{sec:conclude} provides conclusions and perspectives.
We denote vector-valued functions with stylized bold text (i.e., $\boldsymbol
x$), vectors with standard bold text (i.e., $\mathbf{x}$), 
matrices with capital bold letters (i.e., $\mathbf{X} \equiv \begin{bmatrix}
\mathbf{x}_1 & \cdots & \mathbf{x}_r\end{bmatrix}$), and spaces with
calligraphic fonts (i.e., $\mathcal{X}$).
%Model reduction of nonlinear dynamical systems is a growing research area.
%The least-squares Petrov-Galerkin (LSPG) approach has gained popularity as a model reduction technique for nonlinear dynamical systems. Derived at the fully discrete level, the LSPG approach solves a least-squares problem that minimizes the fully discrete residual of the reduced-order model at each time-step. LSPG has been shown to be an effective formulation for nonlinear model reduction on complex problems of interest.
%\begin{comment} 
\section{Mathematical setting}\label{sec:math}
We consider the system of nonlinear ordinary differential equations (ODEs),
\begin{equation}\label{eq:FOM}
 \stateFOMDotArg{t}  = \velocity(\stateFOMArg{}{t},t), \qquad \stateFOM(0) =
	\stateFOMIC,\qquad t \in [0,T]
\end{equation}
where $\stateFOMDot \equiv \frac{d \stateFOM}{dt}$, 
$T \in \mathbb{R}^+$ denotes the final time, $\stateFOM \in \RR{\fomdim}
	\otimes \timeSpace $ with $\stateFOM: [0,T] \rightarrow  \RR{\fomdim}$
	denotes the state implicitly defined as the solution to initial value
	problem~\eqref{eq:FOM}, 
	$\timeSpace$ denotes the set of real-valued functions acting on the time
	domain, 
 $\stateFOMIC \in \mathbb{R}^{\fomdim}$ denotes the initial condition,
	and $\velocity: \mathbb{R}^{\fomdim} \times [0,T] \rightarrow
	\mathbb{R}^{\fomdim}$ with $(\stateyDiscrete,t) \mapsto
	\velocity(\stateyDiscrete,t)$ denotes the velocity, which is possibly
	nonlinear in its first argument. We refer to Eq.~\eqref{eq:FOM} 
	as the ``full-order model" (FOM) ODE.

@@

\subsection{Projection-based reduced-order models}
Projection-based ROMs generate approximate solutions to the FOM
	ODE~\eqref{eq:FOM} by approximating the state in a low-dimensional trial
	subspace. For dynamical 
	systems, two distinct types of space--time trial subspaces are commonly used: 
\begin{enumerate} 
	\item \textit{Space–time trial subspaces with spatial dimension reduction (\spatialAcronym)}. These 
	trial subspaces reduce only the spatial dimension of the FOM, and are commonly used
		in model-reduction methods based on spatial projection followed by time
		integration (e.g., Galerkin projection, LSPG projection.)
	\item \textit{Space–time trial subspaces with spatial and temporal dimension reduction (\spaceTimeAcronym)}. These 
	trial subspaces reduce both the spatial and temporal dimensions of the FOM,
		and are commonly employed 
		employed in model-reduction methods based on space--time projection (e.g.,
		space--time LSPG projection).
\end{enumerate}
 We now describe these two types of trial subspaces and their application to 
	Galerkin, LSPG, and space--time LSPG projection. 
	\KTC{HERE}

\subsection{\spatialAcronym\ trial subspaces}
\spatialAcronym\ trial subspaces approximate the FOM ODE solution
	$\stateFOM$ with an approximation $\approxstate$ that resides in an
	affine space--time trial subspace of dimension $K\ll\fomdim$, i.e., 
\begin{equation}\label{eq:spatial_subspace}
\begin{split}
& \stspaceS \defeq \trialspace \otimes \timeSpace +
	\stateIntercept\otimes\onesFunction,\\ 
& \trialspace \defeq \text{Range}(\basismat),  \\%\label{eq:spatial_subspace2}  \\
&\basismat \defeq \begin{bmatrix}  \basisvec_1  & \cdots &  \basisvec_K \end{bmatrix}, %\label{eq:spatial_subspace3} 
\end{split}
\end{equation}
where 
%	$\timeSpace$ denotes the set of real-valued functions acting on the time
%	domain
%	$[0,T]$, 
$\onesFunction\in\timeSpace$ is defined as $\onesFunction:[0,T]\rightarrow\{1\}$,
	$\basismat \in \RRStar{\romdim}{\fomdim}$ denotes the spatial trial subspace with
	$\RRStar{\romdim}{\fomdim}$ denoting the compact Stiefel manifold (i.e.,  $
	\RRStar{\romdim}{\fomdim}\defeq
	\{ \mathbf{X} \in \RR{\fomdim
	\times \romdim}\, \big|\, \mathbf{X}^T \mathbf{X} = \mathbf{I} \}$),
	$\trialspace\subseteq\RR{\fomdim}$ denotes the spatial trial subspace, and
	$\stateIntercept \in \mathbb{R}^{\fomdim}$ denotes the reference state that
	is often taken to be the initial condition (i.e., $\stateIntercept = \stateFOMIC$). 
%where $\basisvec_i \in \mathbb{R}^N$ are orthonormal basis vectors and $\stateIntercept \in \mathbb{R}^{\fomdim}$ the ``intercept''.
%We refer to $\basismat \in \RRStar{\fomdim \times \romdim}$, with $\RRStar{p \times q}$ denoting the set of orthogonal $q$ frames in $p$, as the spatial trial basis, and $\trialspace$ as the spatial trial space. 
The basis vectors $\basisvec_i$, $i=1,\ldots,K$ are typically constructed
using state snapshots, e.g., via
proper orthogonal decomposition (POD)~\cite{berkooz_turbulence_pod} or the reduced-basis method~\cite{rb_1,rb_2,rb_3,NgocCuong2005,Rozza2008}. 
At any time instance $t\in[0,T]$, ROMs that employ the  \spatialAcronym\ trial subspace approximate the state as
\begin{equation}\label{eq:affine_trialspace}
\stateFOMArg{}{t} \approx \approxstate(t) = \basisspace \genstateArg{}{t} + \stateIntercept,
\end{equation}
where $\genstate \in \RR{\romdim} \otimes \timeSpace$ denotes the generalized
coordinates. Substituting the approximation~\eqref{eq:affine_trialspace} into the FOM ODE~\eqref{eq:FOM} and performing orthogonal
$\elltwo$-projection of the initial condition onto the trial subspace yields
the overdetermined system of ODEs
\begin{equation}\label{eq:g_truncation}
\basisspace \frac{d \genstate} {dt}(t)= \velocity(\basisspace
\genstateArg{}{t} + \stateIntercept,t ), \qquad \genstate(0) = \genstateIC, \qquad t \in [0,T].
\end{equation}
Because Eq.~\eqref{eq:g_truncation} is overdetermined, a solution may not
exist. Typically, either \textit{Galerkin projection} or \textit{least-squares
Petrov--Galerkin projection} is employed to reduce the number of equations
such that a unique solution exists. We now describe these two methods.

@@

\frac{d\genstateGalerkin}{dt}(t)= \massArgnt{n}^{-1} \basisspace^T \stweightingMatArg{n} \velocity(\basisspace
\genstateGalerkinArg{}{t} + \stateIntercept,t), \qquad \genstateGalerkin(0) = \genstateIC, \qquad t \in [0,T],

@@

\frac{d\genstateGalerkin}{dt}(t) = \underset{ \genstateyDiscrete \in \RR{\romdim}

@@

	\velocity(\basisspace \genstateGalerkinArg{}{t} + \stateIntercept,t)
